---
title: "Ospats legacy code"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ospats legacy code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ospats2)
```

# Legacy code, summary

The `ospats2` package includes refactored legacy code from  `Ospats` R-package and the (now deprecated) Julia code.

## `Ospats` R-library

The R-package `Ospats` was found here: https://github.com/brendo1001/ospats/

The original `Ospats::ospatsF` is included as `ospats2::ospatsF` with minor differences to reduce the footprint:

1. The call `fastPdist2(Ar = as.matrix(z), Br = as.matrix(z)))^2` is now  `outer(z, z, "-")^2`
1. The call `fastVsum(Ar = as.matrix(s2))` is now `outer(s2, s2, "+")`
1. The call `fastPdist2(Ar = xy, Br = xy)` is now `as.matrix(dist(xy))`

We rewrote (refactored) `Ospats::ospatsF` for development purposes with these changes:

- only the k-means starting partition algorithm is included (equals parameter `dStart=0`)
- the inputs have been renamed
- the ouputs are simplified
- New option: the covariance matrix can be inputted as-is

The new function is called `ospatsF_ref`, and the renamed parameters are described here:

```{r, eval=FALSE}
                                 # in ospatsF:
ospatsF_ref (x,                  # dat
             covmodel_range,     # dRange
             nstrata,            # dStrata
             niter = 100,        # nCycles
             niter_outer = 100,  # nMaxrun
             verbose = FALSE,    # =
             temperature = 1,    # initialTemperature
             coolingrate = .95,  # coolingRate
             rsquared = 1,       # dRSquare
             Cov                 # not in ospatsF
                                 # Omitted:
                                 # debug, dStart, ClusterStart
)
```

Note that the input for the refactored function should be a tibble, data.frame or similar with named vectors `x,y` for coordinates and `pred,var` for target variable predictions and their variances. For the legacy function `data` should be a matrix with these variables in order `x,y,pred,var`.


Note that the covariance in `ospatsF` is computed as $\sigma_{ij}^2:=0.5(\sigma_i^2+\sigma_j^2)\rho_{ij}$, even though the correct form is $\sigma_{ij}^2=\sigma_i\sigma_j\rho_{ij}$. This can be provided using the `Cov` parameter.


## Julia implementation

The Julia code was found here: https://github.com/jjdegruijter/ospats/

We rewrite the Julia code in the file `ospatsmr`. The functionality is wrapped in a user-facing function, the original code used global variables, see the `main` julia file in the original repository. Note:

- we only implement the partitioning sections of the code, not the sample allocation or sampling sections
- the output is not written into files

The interface aligned with the original R-function:

```{r, eval = FALSE}
ospatsF_julia <- (
                          data,
                          dRange,
                          dRSquare,
                          dMaxrun,
                          nCycles,
                          dStrata,
                          verbose 
                        )
```

And the refactored code with the new inputs:

```{r, eval=FALSE}
ospatsF_julia_ref (
                          x,
                          covmodel_range,
                          rsquared,
                          niter_outer,
                          niter,
                          nstrata,
                          verbose,
                          Cov # optional, new
                        )
```

The Julia implementation differs from the R-implementation in key parts:

- Starting partition: 
  - `ospatsF`: each run starts with the same k-means partition
  - `ospatsF_julia`: each run starts with a completely random allocation
- Realloaction of a unit: The probability of moving a unit from strata to other equals 1 if the objective functions improves
  - `ospatsF`: but slightly worsening move has a small probability too. These probabilities go to zero each iteration (tempering, factor `coolingrate`)
  - `ospatsF_julia`: no tempering, so just consider improvements
  


## An example

We here illustrate how the refactoring does not change the results. Load example data:

```{r}
data("ospats_test_data1", package = "ospats2")
dat1 <- subset(ospats_test_data1, obs)
x <- with(
  dat1,
  data.frame(x = x, y = y, pred = z.pred.mean, var = z.pred.var)
          ) 
dat <- x |> as.matrix()

```

Then run the algorithm from the original R-package `Ospats`. We have duplicated it, with the modification that the C-code was replaced with native R calls (see help for details).

```{r}
## Original:
set.seed(1)
o1 <- ospats2::ospatsF(
  data = dat,
  dRange = 3000,
  dStart = 0,
  dStrata = 3,
  dRSquare = 1,
  dMaxrun = 5,
  nCycles = 100,
  initialTemperature = 1,
  coolingRate = 0.95,
  verbose = FALSE 
)
```

```{r}
# Refactored:
set.seed(1)
o2 <- ospats2::ospatsF_ref(
  x  = x,
  covmodel_range = 3000,
  nstrata = 3,
  rsquared = 1,
  niter_outer = 5,
  niter = 100,
  temperature = 1,
  coolingrate = 0.95,
  verbose = FALSE 
)
```

Check equality.

```{r}
table(old = o1$stratFrame[,5], new= o2$stratification  ) 
# Should be reorder-diagonal

print( c(obj_old = o1$objective, obj_new = o2$Obar ))

```
Pretty much the same. 

Note that since the Julia-based algorithm is different (see above for details), the results might differ. Here they align.

```{r}
# Simpler julia-code algorithm
set.seed(1)
j1 <- ospats2::ospatsF_julia_ref(
  x = x,
  covmodel_range = 3000,
  nstrata = 3,
  niter_outer = 5,
  niter = 100
)

print(table(R=o2$stratification, J=j1$stratification))

print(c(obj_julia=j1$Obar))
```






